Sweat electrolytes and CFTR activity in the era of CFTR modulators

### 2022-10-20 Edith Zemanick/Sweat Test/sweat_test_analysis.Rmd
added tables by specific genotype

```{r setup, include=FALSE}
library(readxl)
library(arsenal)
library(tidyverse)
library(pROC)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "/Users/timvigers/Documents/Work/CF/Edith Zemanick/Sweat Test")
source("~/GitHub/CF/get_genotype_class_severity.R")
```

```{r}
# Import exclusion data from Excel
diagnoses <- read_excel("./Data_Cleaned/AM reclassifications All Patients with CF or Non-CF.xlsx")
exclude <- diagnoses$PatientID[!is.na(diagnoses$Exclude)]
# Import data
df <- read.csv("./Data_Raw/Copy of Sweat Tests_AllFor202558_revised01Feb2021.csv", na.strings = "")
# Remove exclusions from Angela
df <- df[!(df$PatientID %in% exclude), ]
df$Diagnosis <- diagnoses$Diagnosis[match(df$PatientID, diagnoses$PatientID)]
# Add modulator info
cftr <- read.csv("./Data_Raw/Copy of Report CFTR Modulators.csv")
df$Modulator <- apply(df, 1, function(r) {
  mod <- cftr[cftr$Patient.ID == as.numeric(r["PatientID"]), ]
  if (any(duplicated(mod$Age.at.Start..years.))) {
    mod <- mod[-which(duplicated(mod$Age.at.Start..years.)), ]
  }
  mod <- mod[order(mod$Age.at.Start..years.), ]
  age <- as.numeric(r["Age.at.Test.in.Years"])
  as.character(cut(age, c(-Inf, mod$Age.at.Start..years., Inf),
    labels = c("None", mod$Modulator), right = F
  ))
})
# Numeric variables
num_vars = c("Amount","Na","K","Cl")
num_vars = c(paste0(num_vars,"_Left"),paste0(num_vars,"_Right"))
df[,num_vars] = suppressWarnings(lapply(df[,num_vars],as.numeric))
# Filter per Maxie's code
df <- df %>%
  mutate(Age.at.Test.in.Days = Age.at.Test.in.Weeks * 7) %>%
  filter(
    Age.at.Test.in.Years < 50 & Age.at.Test.in.Days >= 2,
    !(Diagnosis == "Non-CF" & PatientID %in%
      c(22, 23, 24, 29, 31, 474, 483, 484, 485, 505, 510, 785, 791, 792, 793)),
    Modulator == "None"
  ) %>%
  distinct(.) %>%
  filter(Sex != "", Demographics.Available == "TRUE")
# Genotype info
df$geno1 <- get_cf_genotype_class_severity(df$Genotypes1)
df$geno2 <- get_cf_genotype_class_severity(df$Genotypes2)
df$geno1[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno2[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno_risk <- paste(df$geno1, df$geno2, sep = "_")
df$geno_risk <- factor(df$geno_risk,
  levels = c(
    "Severe_Severe","Mild_Severe", "Mild_Unknown", "Severe_Mild", "Mild_Mild","Not CF_Not CF",

     "Severe_Unknown", "Unknown_Severe",
    "Unknown_Unknown"
  ),
  labels = c(
    "High","Low", "Low", "Low", "Low", "Not CF",  "Unknown",
    "Unknown", "Unknown"
  )
)
# Which arm?
df$Arm <- apply(df[, c("Cl", "Cl_Left", "Cl_Right", "Na", "Na_Left", "Na_Right", "K", "K_Left", "K_Right")], 1, function(r) {
  k <- sum(!is.na(r))
  if (k == 6) {
    return("Both")
  } else if (k == 3) {
    return("Arm Not Noted")
  } else {
    return(NA)
  }
})

# Difference between arms
df$`Cl Diff. Between Left and Right Arm` = as.numeric(df$Cl_Left) - as.numeric(df$Cl_Right)
df$`Na Diff. Between Left and Right Arm` = as.numeric(df$Na_Left) - as.numeric(df$Na_Right)
# Merge left and right values
df$Amount <- apply(df[, c("Amount", "Amount_Left", "Amount_Right")], 1, function(r) {
  amt <- suppressWarnings(as.numeric(as.character(r)))
  mean(amt, na.rm = T)
})
df$Cl <- apply(df[, c("Cl", "Cl_Left", "Cl_Right")], 1, function(r) {
  cl <- suppressWarnings(as.numeric(as.character(r)))
  mean(cl, na.rm = T)
})
df$Na <- apply(df[, c("Na", "Na_Left", "Na_Right")], 1, function(r) {
  na <- suppressWarnings(as.numeric(as.character(r)))
  mean(na, na.rm = T)
})
df$K <- apply(df[, c("K", "K_Left", "K_Right")], 1, function(r) {
  k <- suppressWarnings(as.numeric(as.character(r)))
  mean(k, na.rm = T)
})
# Sodium to chloride ratio
df$`Na/Cl Ratio` = df$Na /df$Cl
# Sort
df <- df %>% arrange(PatientID, Age.at.Test.in.Weeks)
# Table options (most variables are not normally distributed)
mycontrols  <- tableby.control(numeric.test="kwt",numeric.stats=c("Nmiss", "median", "q1q3"))
labels = list(Age.at.Test.in.Years = "Age (Years)", geno_risk = "Genotype Severity")
```

@@

For those with two sweat electrolyte measures (left and right), the average of the two was used. Each participants' first sweat test value was used for the following tables

@@

```{r results='asis'}
t1_df <- df %>%
  group_by(PatientID) %>%
  filter(row_number() == 1) # First measure for everyone
t1 <- tableby(Diagnosis ~ Sex + Age.at.Test.in.Years + fe(geno_risk) + Amount + Cl + Na + K + `Na/Cl Ratio`,
  data = t1_df,control = mycontrols,simulate.p.value=T
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(Method ~ Sex + Age.at.Test.in.Years + Diagnosis + geno_risk + Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df,control = mycontrols,
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

@@

## Plots

@@

@@

```{r results='asis'}
p <- paired(time ~ Diagnosis +Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,signed.rank.exact = F,
            data = multiples, id = PatientID,control = mycontrols)
summary(p,pfootnote = T,labelTranslations = labels)
```

@@

>>>>

```{r setup, include=FALSE}
library(readxl)
library(arsenal)
library(tidyverse)
library(pROC)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
if(Sys.info()["sysname"] == "Windows"){
  home_dir = "C:/Users/timvigers/OneDrive - The University of Colorado Denver/Vigers/CF/Edith Zemanick/Sweat Test"
  source("C:/Users/timvigers/Documents/GitHub/CF/get_genotype_class_severity.R")
} else if (Sys.info()["sysname"] == "Linux"){
  home_dir = "~/UCD/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Edith Zemanick/Sweat Test"
  source("~/GitHub/CF/get_genotype_class_severity.R")
} else if (Sys.info()["sysname"] == "Darwin"){
  home_dir = "/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Edith Zemanick/Sweat Test"
  source("~/GitHub/CF/get_genotype_class_severity.R")
}
knitr::opts_knit$set(root.dir = home_dir)
rm(home_dir)
```

```{r}
# Import exclusion data from Excel
diagnoses <- read_excel("./Data_Cleaned/AM reclassifications All Patients with CF or Non-CF.xlsx")
exclude <- diagnoses$PatientID[!is.na(diagnoses$Exclude)]
# Import data
df <- read.csv("./Data_Raw/Copy of Sweat Tests_AllFor202558_revised01Feb2021.csv", na.strings = "")
# Remove exclusions from Angela
df <- df[!(df$PatientID %in% exclude), ]
df$Diagnosis <- diagnoses$Diagnosis[match(df$PatientID, diagnoses$PatientID)]
# Add modulator info
cftr <- read.csv("./Data_Raw/Copy of Report CFTR Modulators.csv")
df$Modulator <- apply(df, 1, function(r) {
  mod <- cftr[cftr$Patient.ID == as.numeric(r["PatientID"]), ]
  if (any(duplicated(mod$Age.at.Start..years.))) {
    mod <- mod[-which(duplicated(mod$Age.at.Start..years.)), ]
  }
  mod <- mod[order(mod$Age.at.Start..years.), ]
  age <- as.numeric(r["Age.at.Test.in.Years"])
  as.character(cut(age, c(-Inf, mod$Age.at.Start..years., Inf),
    labels = c("None", mod$Modulator), right = F
  ))
})
# Numeric variables
num_vars = c("Amount","Na","K","Cl")
num_vars = c(paste0(num_vars,"_Left"),paste0(num_vars,"_Right"))
df[,num_vars] = suppressWarnings(lapply(df[,num_vars],as.numeric))
# Filter per Maxie's code
df <- df %>%
  mutate(Age.at.Test.in.Days = Age.at.Test.in.Weeks * 7) %>%
  filter(
    Age.at.Test.in.Years < 50 & Age.at.Test.in.Days >= 2,
    !(Diagnosis == "Non-CF" & PatientID %in%
      c(22, 23, 24, 29, 31, 474, 483, 484, 485, 505, 510, 785, 791, 792, 793)),
    Modulator == "None"
  ) %>%
  distinct(.) %>%
  filter(Sex != "", Demographics.Available == "TRUE")
# Genotype info
## By severity
df$geno1 <- get_cf_genotype_class_severity(df$Genotypes1)
df$geno2 <- get_cf_genotype_class_severity(df$Genotypes2)
df$geno1[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno2[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno_risk <- paste(df$geno1, df$geno2, sep = "_")
df$geno_risk <- factor(df$geno_risk,
  levels = c(
    "Severe_Severe","Mild_Severe", "Mild_Unknown", "Severe_Mild", "Mild_Mild",
    "Not CF_Not CF","Severe_Unknown", "Unknown_Severe","Unknown_Unknown"
  ),
  labels = c(
    "High","Low", "Low", "Low", "Low", "Not CF",  "Unknown",
    "Unknown", "Unknown"
  )
)
## By function
df$geno1_class <- get_cf_genotype_class_severity(df$Genotypes1,out = "class")
df$geno1_class[grep("508",df$Genotypes1)] = "F508del"
df$geno1_class = replace(df$geno1_class,df$geno1_class %in% c("I","II","III"),"MF")
df$geno1_class = replace(df$geno1_class,df$geno1_class %in% c("IV","V"),"RF")

df$geno2_class <- get_cf_genotype_class_severity(df$Genotypes2,out = "class")
df$geno2_class[grep("508",df$Genotypes2)] = "F508del"
df$geno2_class = replace(df$geno2_class,df$geno2_class %in% c("I","II","III"),"MF")
df$geno2_class = replace(df$geno2_class,df$geno2_class %in% c("IV","V"),"RF")

df$geno_class = factor(paste0(pmin(df$geno1_class,df$geno2_class),"/",
                       pmax(df$geno1_class,df$geno2_class)))
# Which arm?
df$Arm <- apply(df[, c("Cl", "Cl_Left", "Cl_Right", "Na", "Na_Left", "Na_Right", "K", "K_Left", "K_Right")], 1, function(r) {
  k <- sum(!is.na(r))
  if (k == 6) {
    return("Both")
  } else if (k == 3) {
    return("Arm Not Noted")
  } else {
    return(NA)
  }
})

# Difference between arms
df$`Cl Diff. Between Left and Right Arm` = as.numeric(df$Cl_Left) - as.numeric(df$Cl_Right)
df$`Na Diff. Between Left and Right Arm` = as.numeric(df$Na_Left) - as.numeric(df$Na_Right)
# Merge left and right values
df$Amount <- apply(df[, c("Amount", "Amount_Left", "Amount_Right")], 1, function(r) {
  amt <- suppressWarnings(as.numeric(as.character(r)))
  mean(amt, na.rm = T)
})
df$Cl <- apply(df[, c("Cl", "Cl_Left", "Cl_Right")], 1, function(r) {
  cl <- suppressWarnings(as.numeric(as.character(r)))
  mean(cl, na.rm = T)
})
df$Na <- apply(df[, c("Na", "Na_Left", "Na_Right")], 1, function(r) {
  na <- suppressWarnings(as.numeric(as.character(r)))
  mean(na, na.rm = T)
})
df$K <- apply(df[, c("K", "K_Left", "K_Right")], 1, function(r) {
  k <- suppressWarnings(as.numeric(as.character(r)))
  mean(k, na.rm = T)
})
# Sodium to chloride ratio
df$`Na/Cl Ratio` = df$Na /df$Cl
# Sort
df <- df %>% arrange(PatientID, Age.at.Test.in.Weeks)
# Table options (most variables are not normally distributed)
mycontrols  <- tableby.control(numeric.test="kwt",numeric.stats=c("Nmiss", "median", "q1q3"))
labels = list(Age.at.Test.in.Years = "Age (Years)", 
              geno_risk = "Genotype Severity",geno_class = "Specific Genotype")
```

@@

For those with two sweat electrolyte measures (left and right), the average of the two was used. Each participants' first sweat test value was used for the following tables.

@@

```{r results='asis'}
t1_df <- df %>%
  group_by(PatientID) %>%
  filter(row_number() == 1) # First measure for everyone
t1 <- tableby(Diagnosis ~ Sex + Age.at.Test.in.Years + fe(geno_risk) + fe(geno_class) + 
                Amount + Cl + Na + K + `Na/Cl Ratio`,
  data = t1_df,control = mycontrols,simulate.p.value = T
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

@@

```{r results='asis'}
t1 <- tableby(Method ~ Sex + Age.at.Test.in.Years + Diagnosis + fe(geno_risk) + fe(geno_class) +
                Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + 
                `Na Diff. Between Left and Right Arm`,
  data = t1_df,control = mycontrols,simulate.p.value = T
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

@@

### Plots

@@

## By specific genotype

```{r results='asis'}
t1 <- tableby(geno_class ~ Amount + Cl + Na + K + `Na/Cl Ratio` + 
                `Cl Diff. Between Left and Right Arm` + 
                `Na Diff. Between Left and Right Arm`,
  data = t1_df,control = mycontrols,
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

## By specific genotype and diagnosis

```{r results='asis'}
t1 <- tableby(geno_class ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Diagnosis == "Cystic Fibrosis",],control = mycontrols
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

```{r results='asis'}
t1 <- tableby(geno_class ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Diagnosis == "CFTR-related disorder",],control = mycontrols
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

```{r results='asis'}
t1 <- tableby(geno_class ~ Amount + Cl + Na + K + `Na/Cl Ratio` + `Cl Diff. Between Left and Right Arm` + `Na Diff. Between Left and Right Arm`,
  data = t1_df[t1_df$Diagnosis == "CRMS",],control = mycontrols
)
summary(t1,pfootnote = T,
  labelTranslations = labels
)
```

### Plots

```{r warning=FALSE}
ggplot(df, aes(x = geno_class, y = Na)) +
  geom_boxplot() +
  theme_bw() + xlab("Specific Genotype") +
  ggtitle("Sweat Sodium")

ggplot(df, aes(x = geno_class, y = K)) +
  geom_boxplot() +
  theme_bw() +
  ggtitle("Sweat Potassium")

ggplot(df, aes(x = geno_class, y = Cl)) +
  geom_boxplot() +
  theme_bw() + xlab("Specific Genotype") +
  ggtitle("Sweat Chloride")

ggplot(df, aes(x = geno_class, y = `Na/Cl Ratio`)) +
  geom_boxplot() +
  theme_bw() + xlab("Specific Genotype") +
  ggtitle("Na/Cl Ratio")
```

@@

```{r results='asis'}
p <- paired(time ~ Diagnosis + Cl + Na + K + `Na/Cl Ratio` + 
              `Cl Diff. Between Left and Right Arm` + 
              `Na Diff. Between Left and Right Arm`,
            signed.rank.exact = F,data = multiples, 
            id = PatientID,control = mycontrols)
summary(p,pfootnote = T,labelTranslations = labels)
```

@@

## By specific genotype

```{r results='asis'}
p <- paired(time ~ Cl + Na + K + `Na/Cl Ratio` + 
              `Cl Diff. Between Left and Right Arm` + 
              `Na Diff. Between Left and Right Arm`,
            signed.rank.exact = F,strata = geno_class,numeric.test = "notest",
            data = multiples, id = PatientID,control = mycontrols)
summary(p,pfootnote = T,labelTranslations = labels)
```

## By genotype class and diagnosis

```{r results='asis'}
p <- paired(time ~ Cl + Na + K + `Na/Cl Ratio`,
            signed.rank.exact = F,strata = geno_class,
            data = multiples %>% filter(Diagnosis == "Cystic Fibrosis"), 
            id = PatientID,control = mycontrols)
summary(p,pfootnote = T,labelTranslations = labels)
```

```{r results='asis'}
p <- paired(time ~ Cl + Na + K + `Na/Cl Ratio`,signed.rank.exact = F,strata = geno_class,
            data = multiples %>% filter(Diagnosis == "Non-CF"), 
            id = PatientID,control = mycontrols)
summary(p,pfootnote = T,labelTranslations = labels)
```
